---
title: "Causal modeling of a dynamic system"
author: "Sara - Yao - Thadryan"
date: "8/6/2019"
output: html_document
---



```{R Libraries, echo = FALSE}
suppressWarnings(suppressMessages(library(gdata)))
suppressMessages(suppressMessages(library(tibble)))
suppressMessages(suppressMessages(library(ode2scm)))
```

It is standard to use mathematic models of how the behavior of a system evolves in time (e.g., ordinary differential equations) to model dynamic systems.  However, modelers are often interested in how an intervention causes a system to behave at equilibrium -- where the system dynamics have stabilized.  For example, in economics, there are differential equations that define the rate of change of price given supply and demand.  However, an economist is generally interested predicting the effect of an intervention on supply (e.g., through a tariff) will affect the price of a product at equilibrium, they are less concerned with predicting exactly how the intervention will cause the price to fluctuate before reaching equilibrium.

## Introduction to the MAPK signaling pathwary

The [mitogen-activated protein kinase](https://en.wikipedia.org/wiki/MAPK/ERK_pathway) (MAPK) pathway is important in many biological processes, such as determination of cell fate. It is a cascade of three proteins, a MAPK (aka Erk), a MAPK kinase (MAP2K, aka Mek), and a MAPK kinase kinase (MAP3K, aka Raf), represented with a causal diagram.
\begin{eqnarray}
& \text{E1} \rightarrow \text{Raf} \rightarrow \text{Mek} \rightarrow \text{Erk} & \label{eq:MAPKdef}
\end{eqnarray}
Here E1 is an input signal to the pathway. The cascade relays the signal from one protein to the next by changing the count of proteins in an active state.

## The biochemical reactions

A protein molecule is in an active state if it has one or more attached phosphoryl groups. Each arrow in \eqref{eq:MAPKdef} combines the reactions of the phosphorylation (i.e., activation) and dephosphorylation (i.e., desactivation). For example, $\text{E1} \rightarrow \text{MAP3K}$ combines two reactions
\begin{eqnarray}
\text{E1} + \text{Raf} \overset{v^{\text{act}}_{K3}}{\rightarrow} \text{E1} + \text{P-Raf} \ \mbox{and}\
\text{P-Raf} \overset{v^{\text{inh}}_{K3}}{\rightarrow} \text{Raf}
\end{eqnarray}

In the first reaction above, a particle of the input signal \text{E1} binds (i.e., activates) a molecule of MAP3K to produce MAP3K with an attached phosphoryl. The rate parameter associated with this reaction is $v^{\text{act}}$. In the second reaction, phosphorylated MAP3K loses its phosphoryl (i.e., deactivates), with the rate $v^{\text{inh}}$.  The remaining arrows between MAP3K and MAP2K, and between MAP2K and MAPK aggregate similar reactions and rate pairs.  However, these reactions represent double-phosphorylations, i.e., they require two activation reactions to activate the protein. This results in a more nuanced dynamics.

## The deterministic mechanistic model

Let $\text{K3}(t)$, $\text{K2}(t)$ and $\text{K}(t)$ denote the counts of fully phosphorylated Raf, Mek, and Erk at time $t$. Since Mek and Erk have to be double phosphorylated to be "on" or "active", then $\text{K2}(t)$ and $\text{K}(t)$ represent counts of double phosphorylated proteins. Let $T_{\text{K3}}$, $T_{\text{K2}}$, and $T_{\text{K}}$ represent the total amount of each of the three proteins (unphosphorylated + single-phosphorylation + double-phosphorylation).  Let $E1$ bet the total amount of input signal, which in this model we assume are constant in time.

We model the system as a set of ordinary differential equations $\mathbb{M}$.

$$
\mathbb{M} =
\left\{\begin{matrix}
 dPRaf : \frac{\mathrm{d} \text{K3}}{\mathrm{d} t} =&  v^{\text{act}}_{K3}\text{E1}(T_{\text{K3}}-\text{K3}(t)) - v^{\text{inh}}_{K3}\text{K3}(t)\\ 
 dPPMek: \frac{\mathrm{d} \text{K2}}{\mathrm{d} t} =&  \frac{(v^{\text{act}}_{K2})^2}{v^{\text{inh}}_{K2}}\text{K3}(t)^2(T_{\text{K2}} - \text{K2}(t)) - v^{\text{act}}_{K2}\text{K3}(t)\text{K2}(t) - v^{\text{inh}}_{K2}\text{K2}(t)\\ 
 dPPErk: \frac{\mathrm{d} \text{K1}}{\mathrm{d} t} =&  \frac{(v^{\text{act}}_{K1})^2}{v^{\text{inh}}_{K1}}\text{K2}(t)^2(T_{\text{K1}} - \text{K1}(t)) - v^{\text{act}}_{K1}\text{K2}(t)\text{K1}(t) - v^{\text{inh}}_{K1}\text{K1}(t)
\end{matrix}\right.
$$

## Apply do operator to the model $\mathbb{M}$

We define an ideal intervention in a dynamic model the same as we would with a structural causal model.  It fixes a protein count variable to a constant value that does not change in time.  For example, the perfect intervention do(K2 = k) applied to $\mathbb{M}$ would modify the model as follows:

$$
\text{do}(\mathbb{M}, K2 := k) = 
\left\{\begin{matrix}
 \frac{\mathrm{d} \text{K3}}{\mathrm{d} t} =&  v^{\text{act}}_{K3}\text{E1}(T_{\text{K3}}-\text{K3}(t)) - v^{\text{inh}}_{K3}\text{K3}(t)\\ 
 \frac{\mathrm{d} \text{K2}}{\mathrm{d} t} =& 0\\ 
 \frac{\mathrm{d} \text{K1}}{\mathrm{d} t} =&  \frac{(v^{\text{act}}_{K1})^2}{v^{\text{inh}}_{K1}}k^2(T_{\text{K1}} - \text{K1}(t)) - v^{\text{act}}_{K1}k\text{K1}(t) - v^{\text{inh}}_{K1}\text{K1}(t)
\end{matrix}\right.
$$

## Define the MAPK model

We define the MAPK model, by assigning a function to each phosphorylated protein. Then we define the initial states, initial rates and a sequence of time.

```{r actual model}
# define MAPK model
dE1 <- function(...) {
  0
}

dPRaf <- function(raf_activate, PRaf, E1, raf_deactivate, TRaf = 100, ...) {
  raf_activate * (TRaf-PRaf) * E1 - raf_deactivate * PRaf
}

dPPMek <- function(mek_activate, mek_deactivate, PRaf, PPMek, TMek = 100, ...) {
  (mek_activate ^ 2) * (PRaf ^ 2) * (TMek - PPMek) / mek_deactivate -       mek_activate * PRaf * PPMek -
    mek_deactivate * PPMek
}

dPPErk <- function(erk_activate, erk_deactivate, PPMek, PPErk, TErk = 100, ...) {
  (erk_activate ^ 2) * (PPMek ^ 2) * (TErk - PPErk) / erk_deactivate -      erk_activate * PPMek * PPErk -
    erk_deactivate * PPErk
}

# define initial rates
rates <- list(
  raf_activate = 0.1,
  raf_deactivate = 0.1,
  mek_activate = 0.1,
  mek_deactivate = 2.0,
  erk_activate = 0.1,
  erk_deactivate = 1.0
)

# initial_states <- list(
#   E1 = 1,
#   PRaf = 5,
#   PPMek = 10,
#   PPErk = 8
# )
initial_states <- list(
  E1 = 1,
  PRaf = 0,
  PPMek = 0,
  PPErk = 0
)

times <- seq(0, 200, by = .1)
```

## A function that applies intervention to the MAPK model

```{R prototype of meta model}
meta_model <- function(list_of_models, list_of_ds, rates, states, interventions = NULL) {
  innerRates <- rates
  innerStates <- states
  innerIntervention <- interventions
      
  # update states with interventions 
  if(!is.null(interventions)) {
    for(int in names(interventions)){
      innerStates[[int]] <- interventions[[int]]
    }
  }
  
  
  transition_function <- function(t, states = innerStates, rates = innerRates, interventions = innerIntervention) {
  # check that list_of_models and list_of_ds are of the same length
  # throw an error if otherwise
    if (length(list_of_models) != length(list_of_ds)) 
      stop("Number of functions do not match nubmer of deltas. Check your input.") 
    
    # update states with interventions 
    if(!is.null(interventions)) {
      for(int in names(interventions)){
        states[[int]] <- interventions[[int]]
      }
    }
    
    # calculate original d
    result <- list()
    var <- as.list(c(rates, states))
    for (i in list_of_models) {
      result <- c(result, (do.call(i, var)))
    }
    
    result <- structure(result, names=list_of_ds)
    
    # update intervention
    for (elem in names(interventions)) {
      result[[paste0("d", elem)]] <- 0
    }
    
    list(unlist(result))  
  }
  
  
  attr(transition_function, 'rates') <- as.list(rates)
  attr(transition_function, 'states') <- as.list(states)
  attr(transition_function, 'interventions') <- as.list(interventions)
  return(transition_function)
}
```

### test meta model function before and after intervention

```{r}
#before intervention
test_bi <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states)
test_bi()

#after intervention
test_ai <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = list(PPMek = 40, PPErk = 30))
test_ai()
```


## ODE Simulation

`ode_sim_meta_model` is just a wrapper for the `deSolve::ode` function for numerically solving  ordinary differential equation functions (ODE's) in R.

```{R simulation}
ode_sim_meta_model <- function(meta_model_object, initial_states, times, interventions = NULL) {
  # interventions <- attr(meta_model_object, "interventions")
  # update states with interventions 
  if(!is.null(interventions)) {
    for(int in names(interventions)){
      initial_states[[int]] <- interventions[[int]]
    }
  }
  initial_states <- structure(as.numeric(initial_states), names = names(initial_states))
  rates <- attr(meta_model_object, 'rates')
  rates <- structure(as.numeric(rates), names = names(rates))
  rates <- as.list(rates)
  as_tibble(
    as.data.frame(
      deSolve::ode(
        y = initial_states,
        times = times,
        func = meta_model_object,
        parms = rates
      )
    )
  )
}
```

### Test ode simulation on a model with and without intervention

```{r}
# simulation on a model without intervention
ode_out_meta_model_bi <- ode_sim_meta_model(test_bi, initial_states, times)
head(ode_out_meta_model_bi)
# simulation on a model with intervention
ode_out_meta_model_ai <- ode_sim_meta_model(test_ai, initial_states, times, interventions = list(PPMek = 40, PPErk = 30))
head(ode_out_meta_model_ai)
```

## Building a deterministic steady state model

Let $\text{equil}()$ be an operator that solves and equilbrates the model.  Then:

$$
\text{equil}(\mathbb{M}) = 
\left\{\begin{matrix}
PRaf: \text{K3}=&  T_{\text{K3}}g_1(\omega_{K3}\text{E1})\\ 
PPMek: \text{K2}=&  T_{\text{K2}}g_2(\omega_{K2}\text{K3}) \\ 
PPErk: \text{K1}=&  T_{\text{K1}}g_2(\omega_{K1}\text{K2})
\end{matrix}\right.
$$

where $\text{K}3$, $\text{K}2$, and $\text{K}1$ are the equilibrium values of $\text{K}3(t)$, $\text{K}2(t)$, and $\text{K}1(t)$, no longer depending on time.  $g_1(u) = \frac{u}{1+u}$, $g_2(u) = \frac{u^2}{1 + u + u^2}$ and $\omega_X = \frac{v^{\text{act}}_{X}}{v^{\text{inh}}_{X}}$.

### The deterministic steady state function

```{R equil model}
mapk_ode_equilM <- function(states, rates, TRaf, TMek, TErk, interventions = NULL) {
  innerState <- states
  innerRate <- rates
  transition_f <- function(states = innerState, rates = innerRate) {

    with(as.list(c(states, rates)), {
      w3 <- raf_activate/raf_deactivate
      w2 <- mek_activate/mek_deactivate
      w1 <- erk_activate/erk_deactivate
      
      t3 <- TRaf
      t2 <- TMek
      t1 <- TErk
      
      if(!is.null(interventions$PRaf)) {
        k3 <- interventions$PRaf
      } else {
        u3 <- w3 * E1
        k3 <- t3 * (u3/(1+u3))      
      }
      
      if(!is.null(interventions$PPMek)) {
        k2 <- interventions$PPMek
      } else {
        u2 <- w2 * k3
        k2 <- t2 * ((u2^2)/(1 + u2 + u2^2))
      }
  
      if(!is.null(interventions$PPErk)) {
        k1 <- interventions$PPErk
      } else {
        u1 <- w1 * k2
        k1 <- t3 * ((u1^2)/(1 + u1 + u1^2))      
      }
  
      list("PRaf" = k3, "PPMek" = k2, "PPErk" = k1)
      
      })
    }
  return(transition_f)
  
}

intervention <- list(PPErk = 30)
# do(equil(M,X=x))
mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention)()
```


## Comparison between steady state and ode simulation without intervention

```{r}
g1 <- function(a) a / (1 + a )
g2 <- function(a) a^2 / (1 + a + a^2)

TRaf = 100
TMek = 100
TErk = 100

E1 <- initial_states$E1
Raf <- TRaf * g1(E1 * rates$raf_activate / rates$raf_deactivate)
Mek <- TMek * g2(Raf * rates$mek_activate / rates$mek_deactivate)
Erk <- TErk * g2(Mek * rates$erk_activate / rates$erk_deactivate)

steady_states <- list(Raf=Raf, Mek=Mek, Erk=Erk)
```

The trajectories (solid lines) represent the evolution of the amounts of MAP3K (green), MAP2K (blue), and MAPK (red) in time. The dashed lines are the calculated equilibrium values.


```{r}
{plot(times, ode_out_meta_model_bi$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_bi$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_bi$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = Raf, col='darkgreen', lty=2)
abline(h = Mek, col='darkblue', lty=2)
abline(h = Erk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```


## Comparison between $\text{equil}(\text{do}(\mathbb{M}, X = x))$ and $\text{do}(\text{equil}(\mathbb{M}, X = x))$

```{r}
# intervene on PRaf
# equil(do(M, X=x))
intervention_PRaf <- list(PRaf = 30)
test_ai_PRaf <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PRaf)

ode_out_meta_model_ai_PRaf <- ode_sim_meta_model(test_ai_PRaf, initial_states, times, interventions = intervention_PRaf)

# do(equil(M, X = x))
do_equil_M_PRaf <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PRaf)()


{plot(times, ode_out_meta_model_ai_PRaf$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PRaf$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PRaf$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PRaf$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PRaf$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PRaf$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```

```{r}
# intervene on PPMek
# equil(do(M, X=x))
intervention_PPMek <- list(PPMek = 40)
test_ai_PPMek <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PPMek)

ode_out_meta_model_ai_PPMek <- ode_sim_meta_model(test_ai_PPMek, initial_states, times, interventions = intervention_PPMek)

# do(equil(M, X = x))
do_equil_M_PPMek <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PPMek)()


{plot(times, ode_out_meta_model_ai_PPMek$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PPMek$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PPMek$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PPMek$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PPMek$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PPMek$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```


```{r}
# intervene on PPErk
# equil(do(M, X=x))
intervention_PPErk <- list(PPErk = 60)
test_ai_PPErk <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PPErk)

ode_out_meta_model_ai_PPErk <- ode_sim_meta_model(test_ai_PPErk, initial_states, times, interventions = intervention_PPErk)

# do(equil(M, X = x))
do_equil_M_PPErk <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PPErk)()


{plot(times, ode_out_meta_model_ai_PPErk$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PPErk$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PPErk$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PPErk$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PPErk$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PPErk$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```

```{r}
# intervene on PRaf and PPMek
# calculate do(equil(M, X = x))

intervention_PRaf_PPMek <- list(PPMek = 40, PRaf = 30)
test_ai_PRaf_PPMek <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PRaf_PPMek)
ode_out_meta_model_ai_PRaf_PPMek <- ode_sim_meta_model(test_ai_PRaf_PPMek, initial_states, times, interventions = intervention_PRaf_PPMek)

do_equil_M_PPMek_PRaf <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PRaf_PPMek)()


{plot(times, ode_out_meta_model_ai_PRaf_PPMek$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PRaf_PPMek$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PRaf_PPMek$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PPMek_PRaf$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PPMek_PRaf$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PPMek_PRaf$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```


```{r}
# intervene on PRaf and PPErk
# equil(do(M, X=x))
intervention_PRaf_PPErk <- list(PRaf = 30, PPErk = 50)
test_ai_PRaf_PPErk <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PRaf_PPErk)

ode_out_meta_model_ai_PRaf_PPErk <- ode_sim_meta_model(test_ai_PRaf_PPErk, initial_states, times, interventions = intervention_PRaf_PPErk)

# do(equil(M, X = x))
do_equil_M_PRaf_PPErk <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PRaf_PPErk)()


{plot(times, ode_out_meta_model_ai_PRaf_PPErk$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PRaf_PPErk$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PRaf_PPErk$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PRaf_PPErk$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PRaf_PPErk$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PRaf_PPErk$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```


```{r}
# intervene on PPMek and PPErk
# equil(do(M, X=x))
intervention_PPMek_PPErk <- list(PPMek = 30, PPErk = 40)
test_ai_PPMek_PPErk <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PPMek_PPErk)

ode_out_meta_model_ai_PPMek_PPErk <- ode_sim_meta_model(test_ai_PPMek_PPErk, initial_states, times, interventions = intervention_PPMek_PPErk)

# do(equil(M, X = x))
do_equil_M_PPMek_PPErk <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PPMek_PPErk)()


{plot(times, ode_out_meta_model_ai_PPMek_PPErk$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PPMek_PPErk$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PPMek_PPErk$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PPMek_PPErk$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PPMek_PPErk$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PPMek_PPErk$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```

```{r }
# intervene on PRaf, PPMek and PPErk b/c why not
# equil(do(M, X=x))
intervention_PRaf_PPMek_PPErk <- list(PRaf = 30, PPMek = 40, PPErk = 50)
test_ai_PRaf_PPMek_PPErk <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk), c("dE1", "dPRaf", "dPPMek", "dPPErk"), rates, initial_states, interventions = intervention_PRaf_PPMek_PPErk)

ode_out_meta_model_ai_PRaf_PPMek_PPErk <- ode_sim_meta_model(test_ai_PRaf_PPMek_PPErk, initial_states, times, interventions = intervention_PRaf_PPMek_PPErk)

# do(equil(M, X = x))
do_equil_M_PRaf_PPMek_PPErk <- mapk_ode_equilM(states = initial_states, rates, TRaf = 100, TMek = 100, TErk = 100, intervention_PRaf_PPMek_PPErk)()


{plot(times, ode_out_meta_model_ai_PRaf_PPMek_PPErk$PRaf, type='l', col='darkgreen', ylim=c(0, 110), ylab = 'amount')
lines(ode_out_meta_model_ai_PRaf_PPMek_PPErk$PPMek, type='l', col='darkblue', ylim=c(0, 110))
lines(ode_out_meta_model_ai_PRaf_PPMek_PPErk$PPErk, type='l', col='darkred', ylim=c(0, 110))
abline(h = do_equil_M_PRaf_PPMek_PPErk$PRaf, col='darkgreen', lty=2)
abline(h = do_equil_M_PRaf_PPMek_PPErk$PPMek, col='darkblue', lty=2)
abline(h = do_equil_M_PRaf_PPMek_PPErk$PPErk, col='darkred', lty=2)
legend(1,114, legend = c("PRaf","PPMek","PPErk"), col = c('darkgreen','darkblue','darkred'), lty = 1)
}
```


## Experiment 1

### do(PRaf = 49.9977)

```{R do(PRaf = 49.9977)}
rates <- list(
  raf_activate = 0.1,
  raf_deactivate = 0.1,
  mek_activate = 0.1,
  mek_deactivate = 2.0,
  erk_activate = 0.1,
  erk_deactivate = 1.0
)

intervention_experiment_1a <- list(PRaf = 49.9977)

test_ai_experiment1a <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk),
                                   c("dE1", "dPRaf", "dPPMek", "dPPErk"),
                                   rates, initial_states,
                                   interventions = intervention_experiment_1a)

out_meta_model_ai_experiment_1a <- ode_sim_meta_model(test_ai_experiment1a,
                                                      initial_states, times,
                                                      interventions = intervention_experiment_1a)
out_meta_model_ai_experiment_1a[nrow(out_meta_model_ai_experiment_1a),]
exp1a_PPErk <- out_meta_model_ai_experiment_1a[nrow(out_meta_model_ai_experiment_1a),"PPErk"]
```






### do(PRaf = 24.9)

```{R do(PRaf = 24.9)}
intervention_experiment_2a <- list(PRaf = 24.9)

test_ai_experiment2a <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk),
                                   c("dE1", "dPRaf", "dPPMek", "dPPErk"),
                                   rates, initial_states,
                                   interventions = intervention_experiment_2a)

out_meta_model_ai_experiment_2a <- ode_sim_meta_model(test_ai_experiment2a,
                                                      initial_states, times,
                                                      interventions = intervention_experiment_2a)
out_meta_model_ai_experiment_2a[nrow(out_meta_model_ai_experiment_2a),]
exp2a_PPErk <- out_meta_model_ai_experiment_2a[nrow(out_meta_model_ai_experiment_2a),"PPErk"]

```

### Causal effect on PPErk

```{r}
exp1a_PPErk - exp2a_PPErk
```

## Experiment 2

### do(PRaf = 40)

```{R do(PRaf = 40)}

rates <- list(
  raf_activate = 0.2,
  raf_deactivate = 0.3,
  mek_activate = 0.2,
  mek_deactivate = 3.0,
  erk_activate = 0.2,
  erk_deactivate = 1.5
)

intervention_experiment_2a <- list(PRaf = 40)

test_ai_experiment2a <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk),
                                   c("dE1", "dPRaf", "dPPMek", "dPPErk"),
                                   rates, initial_states,
                                   interventions = intervention_experiment_2a)

out_meta_model_ai_experiment_2a <- ode_sim_meta_model(test_ai_experiment2a,
                                                      initial_states, times,
                                                      interventions = intervention_experiment_2a)
out_meta_model_ai_experiment_2a[nrow(out_meta_model_ai_experiment_2a),]
exp2a_PPErk <- out_meta_model_ai_experiment_2a[nrow(out_meta_model_ai_experiment_2a),"PPErk"]
```


### do(PRaf = 18.18182)

```{R do(PRaf = 18.18182)}
intervention_experiment_2b <- list(PRaf = 18.18182)

test_ai_experiment2b <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk),
                                   c("dE1", "dPRaf", "dPPMek", "dPPErk"),
                                   rates, initial_states,
                                   interventions = intervention_experiment_2b)

out_meta_model_ai_experiment_2b <- ode_sim_meta_model(test_ai_experiment2b,
                                                      initial_states, times,
                                                      interventions = intervention_experiment_2b)
out_meta_model_ai_experiment_2b[nrow(out_meta_model_ai_experiment_2b),]
exp2b_PPErk <- out_meta_model_ai_experiment_2b[nrow(out_meta_model_ai_experiment_2b),"PPErk"]

```

### Causal effect on PPErk

```{r}
exp2a_PPErk - exp2b_PPErk
```


## Experiment 3

### do(PRaf = 25)

```{R do(PRaf = 25)}
rates <- list(
  raf_activate = 0.1,
  raf_deactivate = 0.3,
  mek_activate = 0.5,
  mek_deactivate = 5.0,
  erk_activate = 0.3,
  erk_deactivate = 4.0
)

intervention_experiment_3a <- list(PRaf = 25)

test_ai_experiment3a <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk),
                                   c("dE1", "dPRaf", "dPPMek", "dPPErk"),
                                   rates, initial_states,
                                   interventions = intervention_experiment_3a)

out_meta_model_ai_experiment_3a <- ode_sim_meta_model(test_ai_experiment3a,
                                                      initial_states, times,
                                                      interventions = intervention_experiment_3a)
out_meta_model_ai_experiment_3a[nrow(out_meta_model_ai_experiment_3a),]
exp3a_PPErk <- out_meta_model_ai_experiment_3a[nrow(out_meta_model_ai_experiment_3a),"PPErk"]
```


### do(PRaf = 9.999999)

```{R do(PRaf = 9.999999)}
intervention_experiment_3b <- list(PRaf = 9.999999)

test_ai_experiment3b <- meta_model(c(dE1, dPRaf, dPPMek, dPPErk),
                                   c("dE1", "dPRaf", "dPPMek", "dPPErk"),
                                   rates, initial_states,
                                   interventions = intervention_experiment_3b)

out_meta_model_ai_experiment_3b <- ode_sim_meta_model(test_ai_experiment3b,
                                                      initial_states, times,
                                                      interventions = intervention_experiment_3b)
out_meta_model_ai_experiment_3b[nrow(out_meta_model_ai_experiment_3b),]
exp3b_PPErk <- out_meta_model_ai_experiment_3b[nrow(out_meta_model_ai_experiment_3b),"PPErk"]

```

### Causal effect on PPErk

```{r}
exp3a_PPErk - exp3b_PPErk
```

### A similar model for stochastic simulation

The stochastic simulation uses the [Gillespie algorithm](https://en.wikipedia.org/wiki/Gillespie_algorithm) to simulate from the MAPK biochemical reaction.  The intuition is this, in both cases each reaction has a rate dependent on the amount of the reaction substrates at a given time point and the reaction rate parameters.  An ODE model treats that rate as determining a continuous change in time.  A stochastic model uses the rate to calculate the probability that a reaction will occur in any given instant.  So the occurance of a reaction is a discrete event, thus the trajectories make small random jumps. To learn more about this, [here](https://d1b10bmlvqabco.cloudfront.net/attach/jv2j4bw56an62b/hjrof4swrey3fd/jz3da8jk1uw7/lecture6.pdf) are some lovely slides.  

## Modify the SDE simulation function by adding intervention as the input

```{r stochastic}
# initial states are different now because we add Raf, Mek, PMek, Erk and PErk as well
initial_states <-  list(E1=1, Raf=100, PRaf=0, Mek=100, PMek=0, PPMek=0, Erk=100, PErk=0, PPErk=0)

mapk_sde <- function(states, rates, interventions = NULL){
  sde <- list()
  
  sde$Pre <- matrix(
    c(
      1, 1, 0, 0, 0, 0, 0, 0, 0,
      0, 0, 1, 0, 0, 0, 0, 0, 0,
      0, 0, 1, 1, 0, 0, 0, 0, 0,
      0, 0, 0, 0, 1, 0, 0, 0, 0,
      0, 0, 1, 0, 1, 0, 0, 0, 0,
      0, 0, 0, 0, 0, 1, 0, 0, 0,
      0, 0, 0, 0, 0, 0, 1, 0, 0,
      0, 0, 0, 0, 0, 0, 0, 1, 0,
      0, 0, 0, 0, 0, 0, 0, 1, 0,
      0, 0, 0, 0, 0, 0, 0, 0, 1
    ), nrow=10, ncol=9, byrow=T
  )
  colnames(sde$Pre) <- c("E1","Raf", "PRaf", "Mek", "PMek", "PPMek", "Erk", "PErk", "PPErk")
  sde$Post <- matrix(
    c(
      1, 0, 1, 0, 0, 0, 0, 0, 0,
      0, 1, 0, 0, 0, 0, 0, 0, 0,
      0, 0, 1, 0, 1, 0, 0, 0, 0,
      0, 0, 0, 1, 0, 0, 0, 0, 0,
      0, 0, 1, 0, 0, 1, 0, 0, 0,
      0, 0, 0, 0, 1, 0, 0, 0, 0,
      0, 0, 0, 0, 0, 0, 0, 1, 0,
      0, 0, 0, 0, 0, 0, 1, 0, 0,
      0, 0, 0, 0, 0, 0, 0, 0, 1,
      0, 0, 0, 0, 0, 0, 0, 1, 0
    ), nrow=10, ncol=9, byrow=T
  )
  colnames(sde$Post) <- c("E1","Raf", "PRaf", "Mek", "PMek", "PPMek", "Erk", "PErk", "PPErk")
  innerIntervention <- interventions
  
  sde$h <- function(states, t, parameters=rates, interventions = innerIntervention){
    # update the initial states
    if(!is.null(interventions)) {
      for(int in names(interventions)){
        states[[int]] <- interventions[[int]]
      }
    }
    with(as.list(c(states, parameters, interventions)), {
      if(!is.null(interventions)) {
        for(int in names(interventions)){
          sde$Pre[,int] <- 0
          sde$Post[,int] <- 0
        }
      }
    
    out <- c(
      raf_activate * Raf * E1,
      raf_deactivate * PRaf,
      mek_activate * PRaf * Mek,
      mek_deactivate * PMek,
      mek_activate * PRaf * PMek,
      mek_deactivate * PPMek,
      erk_activate * PPMek * Erk,
      erk_deactivate * PErk,
      erk_activate * PPMek * PErk,
      erk_deactivate * PPErk
    )
    
    names(out) <- c("RafToPRaf","PRafToRaf",
                    "MekToPMek","PMekToMek","PMekToPPMek","PPMekToPMek",
                    "ErkToPErk","PErkToErk","PErkToPPErk","PPErkToPErk")
    if(!is.null(interventions)) {
      for(int in names(interventions)){
        out[which(grepl(int, names(out), ignore.case = TRUE) == TRUE)] <- 0
      }
    }
      out <- unname(out)
      
      return(out)
    })
  }
  transition_function <- StepGillespie(sde)
  return(transition_function)
}
```


### Modify sde_sim function by adding intervention as the input 

```{r}
sde_sim <- function(transition_function, initial_states, times, interventions = NULL) {
  if(!is.null(interventions)) {
    for(int in names(interventions)){
      initial_states[[int]] <- interventions[[int]]
    }
  }
  initial_states <- structure(as.numeric(initial_states), names = names(initial_states))
  t_delta <- times[2] - times[1]
  out <- as_tibble(
    smfsb::simTs(initial_states, times[1], times[length(times)], t_delta, transition_function)
  )
  out$time <- times
  out <- out[, c('time', setdiff(names(out), 'time'))]
  return(out)
}
# transition_function <- stoc_transition_func
# interventions <- intervention_raf
# sde_out <- sde_sim(stoc_transition_func, initial_states, times,intervention_raf)
```


### Intervene on Raf
```{r}
times <- seq(0, 0.5, by = .1)
intervention_raf <- list(Raf = 70, PRaf = 30)
faster_rates <- lapply(rates, `*`, 20)
stoc_transition_func <- mapk_sde(initial_states, faster_rates,intervention_raf)
sde_out <- sde_sim(stoc_transition_func, initial_states, times,intervention_raf)
sde_out[nrow(sde_out),]

```

### Intervene on Mek
```{r}
times <- seq(0, 0.5, by = .1)
intervention_mek <- list(Mek=50, PMek=20, PPMek=30)
faster_rates <- lapply(rates, `*`, 20)
stoc_transition_func <- mapk_sde(initial_states, faster_rates,intervention_mek)
sde_out <- sde_sim(stoc_transition_func, initial_states, times,intervention_mek)
sde_out[nrow(sde_out),]

```

### Intervene on Erk
```{r}
times <- seq(0, 0.5, by = .1)
intervention_erk <- list(Erk=50, PErk=20, PPErk=30)
faster_rates <- lapply(rates, `*`, 20)
stoc_transition_func <- mapk_sde(initial_states, faster_rates,intervention_erk)
sde_out <- sde_sim(stoc_transition_func, initial_states, times,intervention_erk)
sde_out[nrow(sde_out),]

```

### Intervene on Raf and Mek
```{r}
times <- seq(0, 0.5, by = .1)
intervention_mek_raf <- list(Raf = 70, PRaf = 30,Mek=50, PMek=20, PPMek=30)
faster_rates <- lapply(rates, `*`, 20)
stoc_transition_func <- mapk_sde(initial_states, faster_rates,intervention_mek_raf)
sde_out <- sde_sim(stoc_transition_func, initial_states, times,intervention_mek_raf)
sde_out[nrow(sde_out),]

```

